---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 6,
  fig.height = 6,
  dpi = 96
)
```

# enrichVolcano

> Draw a volcano-in-ring composite from any DA table + any enrichment
> table you've already computed.

`enrichVolcano` is a plotting-only R package. It does not run
differential abundance, it does not run enrichment, and it does not ship
gene-set databases. Compute those upstream with the tool that fits your
study — `fgsea`, `clusterProfiler`, `enrichR`, anything that returns a
tidy frame — and pass the two tables to `volcano_ring()`.

## What it draws

- the differential-abundance volcano in the centre
- one NES-coloured arc per enriched pathway in a ring around the volcano
- tick lines from each pathway arc to its leading-edge genes inside the volcano
- a multi-contrast grid composer (`volcano_ring_grid()`) for side-by-side panels

## Install

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("Dustyn-T-Lewis/enrichVolcano")
```

## Minimum example

```{r example, eval = FALSE}
library(enrichVolcano)

da <- read.csv(system.file("extdata", "examples", "yvo_da.csv",
  package = "enrichVolcano"
))
en <- read.csv(system.file("extdata", "examples", "yvo_enrichment.csv",
  package = "enrichVolcano"
))

ctr <- "Training_Young"
da1 <- da[da$contrast == ctr, ]
en1 <- en[en$contrast == ctr, ]
names(da1)[names(da1) == "adj.P.Val"] <- "padj"

volcano_ring(da1, en1, title = "YvO", subtitle = ctr)
```

See `vignette("getting-started")` for the full walkthrough and
`vignette("input-contract")` for the column conventions plus a
limma / DESeq2 / edgeR / DEP / fgsea / clusterProfiler / enrichR
cross-walk.

## Public API (v0.3.0)

| Function | Role |
|---|---|
| `volcano_ring()` | one composite for one contrast — returns a `ggplot` |
| `volcano_ring_grid()` | a grid of composites, one per contrast |
| `volcano_ring_theme()` | theme + palette factory (`default`, `viridis`, `okabe`) |
| `print.volcano_ring_grid()` | print method for the grid object |

## Upstream tools we consume

`enrichVolcano` is a thin wrapper around outputs from existing enrichment
tooling. When you cite the figure, the methodological credit belongs
upstream:

- `fgsea` — Korotkevich G et al. 2021, DOI 10.1101/060012
- `clusterProfiler` — Wu T et al. 2021 *Innovation*
- `enrichR` — Chen EY et al. 2013 *BMC Bioinformatics*; Kuleshov MV et al. 2016 *Nucleic Acids Res*

Useful references when choosing what to run and how to interpret it:

- Xiao Y et al. 2014 *Bioinformatics* 30(6):801-7, PMID 22321699 — π-value
- Timmons JA et al. 2015 *Genome Biol* 16:186, PMID 26346307 — ORA universe
- Reimand J et al. 2019 *Nat Protoc* 14:482-517, PMID 30664679 — protocol
- Wijesooriya K et al. 2022 *PLoS Comput Biol* 18:e1009935, PMID 35263338 — background bias

## Versioning

v0.3.0 is a breaking release: the v0.2.x enrichment engine, the Shiny
app, and the standalone volcano / ring functions are gone. v0.2.0 stays
installable from the `v0.2.0` tag if you need the old hero
`enrich_volcano()` wrapper or the GUI.

The Shiny app moves to a sibling repo (`enrichVolcanoApp`) and depends
on `enrichVolcano >= 0.3.0`.

## License

MIT. See `LICENSE`.
