---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# enrichVolcano

<!-- badges: start -->
[![R-CMD-check](https://github.com/Dustyn-T-Lewis/enrichVolcano/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Dustyn-T-Lewis/enrichVolcano/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

## Overview

enrichVolcano builds composite volcano + enrichment ring plots from differential abundance results. Volcano on top, enrichment terms as labeled arcs underneath, both stitched as a single annotated patchwork. Returns a ggplot-compatible object that `ggsave()` saves directly.

The package accepts ten common input formats (tidy, wide-suffix, DEP, proteoDA, limma, DESeq2, edgeR, MSstats, proDA, DEqMS, MaxQuant, Perseus) via `ev_validate()`, supports two pi-score variants (Xiao 2014 Eq.1 and Eq.2), four p-adjustment methods (BH, Bonferroni, q-value, IHW), and 21 registered gene-set databases across seven species.

## Installation

```r
# Development version
remotes::install_github("Dustyn-T-Lewis/enrichVolcano")
```

## A 30-second example

```{r example, eval = FALSE}
library(enrichVolcano)

# Load a bundled YvO subsample
data <- readRDS(system.file("extdata", "examples", "yvo_tidy.rds",
                            package = "enrichVolcano"))

# Mock pathway sets (skip msigdbr in this demo)
paths <- list(
  UP_GENES = head(data$gene[data$logFC > 0], 30),
  DN_GENES = head(data$gene[data$logFC < 0], 30)
)

p <- enrich_volcano(
  data, contrast = "Aging",
  databases = list(test = paths),
  p_method = "pi_eq2", p_adjust = "BH",
  enrich_padj = 0.5
)

p
```

The output prints the reconstructible call plus a one-line data summary, then renders the patchwork. Both panels share the contrast title.

## Why a composed plot

Volcano panels show per-protein effects. Enrichment plots show per-pathway aggregates. Drawing them separately loses the connection between a hit and the pathway it drives. The composite preserves the hit-to-pathway lineage and produces one figure per contrast instead of two.

## Reproducibility

Every output carries two attributes:

- `attr(p, "ev_call")` — `match.call()` for exact reconstruction
- `attr(p, "ev_data")` — list of intermediate tibbles: validated input, pi-scores, enrichment, dedup result

These let you rebuild any panel without re-running the pipeline.

## Pipeline functions

`enrich_volcano()` orchestrates eight composable steps, all exported:

```r
ev_validate()   # input adapter (10 sources)
pi_score()   # Xiao 2014 Eq.1 or Eq.2
adjust_p()   # BH / Bonferroni / qvalue / IHW
ev_enrich()     # fgsea + ORA
ev_collapse()   # Jaccard pathway dedup
ev_volcano()    # volcano-only ggplot
ring_plot()       # enrichment ring ggplot
ev_compose()    # patchwork composition
```

Plus `list_databases()` / `database_info()` for the registry and `ev_theme()` for styling.

## Learn more

- `vignette("enrichVolcano")` — 30-second example + pipeline anatomy
- `vignette("scoring")` — pi-score variants, p-value adjustment methods
- `vignette("pathway-dedup")` — Jaccard vs collapsePathways, scope choices
- `vignette("databases")` — 21 databases compared, decision flowchart
- `vignette("customising")` — themes, palettes, multi-contrast layouts
- `vignette("enrichVolcano-faq")` — common questions

## Citation

```r
citation("enrichVolcano")
```

The package implements the pi-score from Xiao Y et al. (2014) *Bioinformatics* 30(6):801, fgsea from Korotkevich G et al. (2021) bioRxiv 060012, and Benjamini-Hochberg FDR from Benjamini & Hochberg (1995) JRSS B 57(1):289.

## License

MIT
