---
title: "Customising enrichVolcano plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customising enrichVolcano plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(enrichVolcano)
```

## Themes and palettes

`ev_theme()` returns a list with three elements: `base_size`, `theme`
(a ggplot2 theme object), and `palette` (named hex codes).

```{r theme-default}
th <- ev_theme()
names(th)
th$palette
```

The default palette is locked. `up = "#D6604D"` is the rust-red used
across the YvO and CvH figures. `down = "#4393C3"` is the
complementary blue. `ns = "grey70"` is the non-significant point
colour. `nes_scale = c("#08306B", "white", "#67000D")` is the
diverging colour ramp used for NES on the ring panel.

For poster fonts, pass a larger base size:

```{r theme-poster, eval = FALSE}
th <- ev_theme(base_size = 14)
```

The `theme` element drops into any ggplot via `+ th$theme`. To
change colours, edit the returned palette element and pass the
whole list to the volcano or ring functions via their `theme`
argument:

```{r theme-custom, eval = FALSE}
th <- ev_theme()
th$palette$up   <- "#B2182B"
th$palette$down <- "#2166AC"

p_volcano <- ev_volcano(yvo, contrast = "Aging", theme = th)
```

`ev_theme()$palette` is a plain named list, so any standard list
edit works. Both `ev_volcano()` and `ring_plot()` read the palette out
of `theme$palette` at draw time.

## Volcano label control

The volcano panel honours two label arguments passed through the
`volcano` list to `enrich_volcano()`.

```{r volcano-labels, eval = FALSE}
p <- enrich_volcano(
  data     = yvo,
  contrast = "Aging",
  volcano  = list(
    label_n  = 10,
    label_by = "pi_eq2"
  )
)
```

`label_n` is the number of top genes to label. `label_by` is the
column used to rank them: `"pi_eq2"`, `"pi_eq1"`, `"adj.P.Val"`, or
`"logFC"`. Smaller is better for pi-score and p-value columns;
absolute value is used for `logFC`.

To force-label a specific set of genes regardless of rank, call
`ev_volcano()` directly with `label_genes`:

```{r volcano-force, eval = FALSE}
p_volcano <- ev_volcano(
  yvo,
  contrast    = "Aging",
  label_n     = 10,
  label_by    = "pi_eq2",
  label_genes = c("MYH7", "ATP5F1A", "PPARGC1A")
)
```

`enrich_volcano()` in v0.1 forwards `label_n` and `label_by` to
`ev_volcano()` but does not pass `label_genes` through the `volcano`
list. Forwarding is planned for v0.2. Until then, build the volcano
panel with `ev_volcano()`, the ring with `ring_plot()`, and stitch
with `ev_compose()`.

`ggrepel` parameters are set inside `ev_volcano()`:
`max.overlaps = 50`, `force = 6`, `box.padding = 0.3`. These match
the YvO 2025 figure series. They are not user-facing in v0.1; if you
need different repel behaviour, call `ev_volcano()` directly and add
your own `geom_text_repel()` layer:

```{r volcano-direct, eval = FALSE}
p <- ev_volcano(yvo, contrast = "Aging")
p <- p + ggrepel::geom_text_repel(
  data = subset(yvo, gene %in% c("MYH7", "ATP5F1A")),
  aes(label = gene),
  max.overlaps = Inf
)
```

The volcano returned by `ev_volcano()` is a plain ggplot, so any
ggplot2 modifier works.

## Ring encodings

The ring panel maps three pathway-level columns to three visual
channels. All three are exposed through the `ring` list.

```{r ring-args, eval = FALSE}
p <- enrich_volcano(
  data     = yvo,
  contrast = "Aging",
  ring     = list(
    max_terms = 8,
    order_by  = "padj",
    order_dir = "asc",
    magnitude = "neg_log_padj",
    color     = "nes"
  )
)
```

`magnitude` sets arc thickness. Options are `"neg_log_padj"`
(default; thicker = more significant), `"nes"` (thicker = stronger
enrichment regardless of direction), and `"gene_count"` (thicker =
larger pathway).

`color` sets arc fill. Options are `"nes"` (default; diverging blue
to white to red across the NES range), `"neg_log_padj"` (sequential
ramp), and `"direction"` (categorical: up red, down blue).

`order_by` controls the arc layout around the ring. Options are
`"padj"` (default), `"size"`, `"NES"`, and `"alphabetic"`. Combine
with `order_dir = "asc"` or `"desc"` to flip direction.

A practical example: map arc thickness to gene count and arc fill to
NES, ordered by absolute NES:

```{r ring-variant, eval = FALSE}
p <- enrich_volcano(
  data     = yvo,
  contrast = "Aging",
  ring     = list(
    max_terms = 8,
    order_by  = "NES",
    order_dir = "desc",
    magnitude = "gene_count",
    color     = "nes"
  )
)
```

This layout shows the biggest pathways first and uses fill for
direction, which works well when reviewers want to see effect size
at a glance.

## Multi-contrast layouts

Pass a vector to `contrast` to build one volcano-ring pair per
contrast. `ev_compose()` arranges the pairs into a single patchwork.

```{r multi-contrast, eval = FALSE}
p <- enrich_volcano(
  data     = yvo,
  contrast = c("Aging", "Training_Young", "Training_Old",
               "Interaction"),
  databases = list(mock = mock_msigdb),
  facet    = list(nrow = 2, ncol = 2)
)
```

`facet$nrow` and `facet$ncol` go to `patchwork::plot_layout()`. The
result is a 2x2 grid of volcano-ring pairs. Each pair keeps its own
contrast title; the outer layout collects guides into a single
legend strip via `plot_layout(guides = "collect")`.

The 4-contrast YvO example produces the same composite that appears
in F02 of the YvO 2025 pipeline. The fixture has the four contrasts
needed; substitute a real database for `mock_msigdb` to reproduce
the published figure.

## Term-label cleaning

Ring labels can be long. `HALLMARK_OXIDATIVE_PHOSPHORYLATION` is 35
characters and crowds the ring. `ev_default_clean_label()` strips
the database prefix (`HALLMARK_`, `REACTOME_`, `KEGG_LEGACY_`,
`KEGG_MEDICUS_`) and replaces underscores with spaces, leaving
`Oxidative Phosphorylation`.

The default fires automatically inside `ring_plot()`. For v0.1, the
clean function is fixed. A custom override
(`ring$label_clean_fn = function(x) gsub(...)`) is planned for v0.2.
Until then, post-process the ggplot returned by `ring_plot()`:

```{r label-clean, eval = FALSE}
p <- ring_plot(enr, contrast = "Aging")
p$data$label <- gsub("Oxidative", "OxPhos", p$data$label)
```

This works because the label data lives on the ggplot, not in a
separate text annotation.

## Exporting for journals

`enrich_volcano()` returns a patchwork. `ggsave()` works directly:

```{r ggsave-pdf, eval = FALSE}
ggplot2::ggsave("fig.pdf", p, width = 180, height = 120, units = "mm")
```

A sizing gotcha: nested patchwork annotations need
`plot_layout(guides = "collect")` at the outer level to share
legends. `ev_compose()` already sets this, so multi-contrast figures
collect guides automatically. If you wrap an `enrich_volcano()`
result inside a larger patchwork, set `guides = "collect"` again at
the new outer level or the legends will multiply.

Common sizes for journal submission:

* Single column: `width = 89, height = 80, units = "mm"`.
* One and a half column: `width = 120, height = 90, units = "mm"`.
* Double column: `width = 180, height = 120, units = "mm"`.

For PNG at 300 dpi:

```{r ggsave-png, eval = FALSE}
ggplot2::ggsave("fig.png", p, width = 180, height = 120,
                units = "mm", dpi = 300)
```

For TIFF with LZW compression (some journals require it):

```{r ggsave-tiff, eval = FALSE}
ggplot2::ggsave("fig.tiff", p, width = 180, height = 120,
                units = "mm", dpi = 300, compression = "lzw")
```

If you hit issues, check `vignette("enrichVolcano-faq")`.
