---
title: "Customising enrichVolcano plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customising enrichVolcano plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(enrichVolcano)
```

## Themes and palettes

`ev_theme()` returns a list with three elements: `base_size`, `theme`
(a ggplot2 theme object), and `palette` (named hex codes).

```{r theme-default}
th <- ev_theme()
names(th)
th$palette
```

The default palette is locked. `up = "#D6604D"` is the rust-red used
across the YvO and CvH figures. `down = "#4393C3"` is the
complementary blue. `ns = "grey70"` is the non-significant point
colour. `nes_scale = c("#08306B", "white", "#67000D")` is the
diverging colour ramp used for NES on the ring panel.

For poster fonts, pass a larger base size:

```{r theme-poster, eval = FALSE}
th <- ev_theme(base_size = 14)
```

The `theme` element drops into any ggplot via `+ th$theme`. To
change colours, edit the returned palette element and pass the
whole list to the volcano or ring functions via their `theme`
argument:

```{r theme-custom, eval = FALSE}
th <- ev_theme()
th$palette$up   <- "#B2182B"
th$palette$down <- "#2166AC"

p_volcano <- ev_volcano(yvo, contrast = "Aging", theme = th)
```

`ev_theme()$palette` is a plain named list, so any standard list
edit works. Both `ev_volcano()` and `ring_plot()` read the palette out
of `theme$palette` at draw time.

## Two ways to plot

`enrich_volcano()` is the opinionated hero: it returns the canonical
volcano-in-ring composite with fixed encodings (up/down counts on the
volcano, arc height = `-log10(padj)`, fill = NES). It is tuned through
`p_threshold`, `enrich_padj`, `dedup`, `ring$max_terms`, `facet`, and
`theme`.

When you need per-gene labels or different ring encodings, drop down to
the standalone trio the package also exports — `ev_volcano()`,
`ring_plot()`, and `ev_compose()` — and assemble the panels yourself.
The rest of this vignette uses those functions for the tunable options.

## Volcano label control

The composite volcano reports up/down counts rather than labelling
genes. For a labelled volcano, call `ev_volcano()` directly. `label_n`
is the number of top genes to label and `label_by` is the ranking
column (`"pi_eq2"`, `"pi_eq1"`, `"adj.P.Val"`, or `"logFC"`; smaller is
better for pi-score and p-value columns, absolute value for `logFC`).

```{r volcano-labels, eval = FALSE}
p_volcano <- ev_volcano(
  yvo,
  contrast = "Aging",
  label_n  = 10,
  label_by = "pi_eq2"
)
```

To force-label a specific set of genes regardless of rank, add
`label_genes`:

```{r volcano-force, eval = FALSE}
p_volcano <- ev_volcano(
  yvo,
  contrast    = "Aging",
  label_n     = 10,
  label_by    = "pi_eq2",
  label_genes = c("MYH7", "ATP5F1A", "PPARGC1A")
)
```

Build the matching ring with `ring_plot()` and stitch the two with
`ev_compose()` to get a labelled composite.

`ggrepel` parameters are set inside `ev_volcano()`:
`max.overlaps = 50`, `force = 6`, `box.padding = 0.3`. These match
the YvO 2025 figure series. They are not user-facing in v0.1; if you
need different repel behaviour, call `ev_volcano()` directly and add
your own `geom_text_repel()` layer:

```{r volcano-direct, eval = FALSE}
p <- ev_volcano(yvo, contrast = "Aging")
p <- p + ggrepel::geom_text_repel(
  data = subset(yvo, gene %in% c("MYH7", "ATP5F1A")),
  aes(label = gene),
  max.overlaps = Inf
)
```

The volcano returned by `ev_volcano()` is a plain ggplot, so any
ggplot2 modifier works.

## Ring encodings

In the hero composite the ring encoding is fixed (arc height =
`-log10(padj)`, fill = NES) and only `ring$max_terms` is tunable. To
remap the visual channels, draw the ring with the standalone
`ring_plot()`, which exposes three encodings. Start from an enrichment
table:

```{r ring-enrich, eval = FALSE}
enr <- ev_enrich(yvo, contrast = "Aging",
                 databases = list(mock = mock_msigdb), enrich_mode = "fgsea")
```

```{r ring-args, eval = FALSE}
p_ring <- ring_plot(
  enr, contrast = "Aging",
  max_terms = 8,
  order_by  = "padj",
  order_dir = "asc",
  magnitude = "neg_log_padj",
  color     = "nes"
)
```

`magnitude` sets arc thickness. Options are `"neg_log_padj"`
(default; thicker = more significant), `"nes"` (thicker = stronger
enrichment regardless of direction), and `"gene_count"` (thicker =
larger pathway).

`color` sets arc fill. Options are `"nes"` (default; diverging blue
to white to red across the NES range), `"neg_log_padj"` (sequential
ramp), and `"direction"` (categorical: up red, down blue).

`order_by` controls the arc layout around the ring. Options are
`"padj"` (default), `"size"`, `"NES"`, and `"alphabetic"`. Combine
with `order_dir = "asc"` or `"desc"` to flip direction.

A practical example: map arc thickness to gene count and arc fill to
NES, ordered by descending NES:

```{r ring-variant, eval = FALSE}
p_ring <- ring_plot(
  enr, contrast = "Aging",
  max_terms = 8,
  order_by  = "NES",
  order_dir = "desc",
  magnitude = "gene_count",
  color     = "nes"
)
```

This layout shows the biggest pathways first, which works well when
reviewers want to see effect size at a glance.

## Multi-contrast layouts

Pass a vector to `contrast` to build one volcano-ring pair per
contrast. `patchwork::wrap_plots()` arranges the pairs into a single
figure.

```{r multi-contrast, eval = FALSE}
p <- enrich_volcano(
  data     = yvo,
  contrast = c("Aging", "Training_Young", "Training_Old",
               "Interaction"),
  databases = list(mock = mock_msigdb),
  facet    = list(nrow = 2, ncol = 2)
)
```

`facet$nrow` and `facet$ncol` go to `patchwork::wrap_plots()`. The
result is a 2x2 grid of volcano-ring pairs, each keeping its own
contrast title.

The 4-contrast YvO example mirrors the composite that appears in F03
of the YvO 2025 pipeline. The fixture has the four contrasts needed;
substitute a real database for `mock_msigdb` to reproduce the
published figure.

## Term-label cleaning

Ring labels can be long. `HALLMARK_OXIDATIVE_PHOSPHORYLATION` is 35
characters and crowds the ring. By default `ring_plot()` strips the
database prefix (`HALLMARK_`, `REACTOME_`, `KEGG_LEGACY_`,
`KEGG_MEDICUS_`) and replaces underscores with spaces, leaving
`Oxidative Phosphorylation`. (The hero composite uses a richer cleaner
that also expands common acronyms and wraps long names.)

`ring_plot()` accepts a `label_clean_fn` argument, so you can supply
your own naming function instead of the default:

```{r label-clean, eval = FALSE}
p_ring <- ring_plot(
  enr, contrast = "Aging",
  label_clean_fn = function(x) gsub("Oxidative", "OxPhos", x)
)
```

This works because the label data lives on the ggplot, not in a
separate text annotation.

## Exporting for journals

`enrich_volcano()` returns a patchwork. `ggsave()` works directly:

```{r ggsave-pdf, eval = FALSE}
ggplot2::ggsave("fig.pdf", p, width = 180, height = 120, units = "mm")
```

A sizing gotcha: each contrast panel carries its own NES colourbar, so
a multi-contrast figure shows one legend per panel. To share a single
legend, collect guides at the outer level:

```{r collect-guides, eval = FALSE}
p & patchwork::plot_layout(guides = "collect")
```

The same applies if you wrap an `enrich_volcano()` result inside a
larger patchwork: set `guides = "collect"` at the new outer level or
the legends will multiply.

Common sizes for journal submission:

* Single column: `width = 89, height = 80, units = "mm"`.
* One and a half column: `width = 120, height = 90, units = "mm"`.
* Double column: `width = 180, height = 120, units = "mm"`.

For PNG at 300 dpi:

```{r ggsave-png, eval = FALSE}
ggplot2::ggsave("fig.png", p, width = 180, height = 120,
                units = "mm", dpi = 300)
```

For TIFF with LZW compression (some journals require it):

```{r ggsave-tiff, eval = FALSE}
ggplot2::ggsave("fig.tiff", p, width = 180, height = 120,
                units = "mm", dpi = 300, compression = "lzw")
```

If you hit issues, check `vignette("enrichVolcano-faq")`.
