---
title: "Customizing the composite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing the composite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  dpi = 96,
  message = FALSE
)
```

Every composite comes from one call to `volcano_ring()`. The defaults are
chosen to read well, and you reach everything else through two places:

- **`volcano_ring_theme()`** controls the *looks*: point colours and the
  NES ramp.
- **`volcano_ring()` arguments** control the *data mapping and layout*:
  what counts as significant, how arcs are ordered and sized, what gets
  labelled.

This vignette walks the knobs section by section on the bundled YvO
example.

```{r data}
library(enrichVolcano)

da <- read.csv(system.file("extdata", "examples", "yvo_da.csv",
  package = "enrichVolcano"
))
en <- read.csv(system.file("extdata", "examples", "yvo_enrichment.csv",
  package = "enrichVolcano"
))
ctr <- "Training_Young"
da1 <- da[da$contrast == ctr, , drop = FALSE]
names(da1)[names(da1) == "adj.P.Val"] <- "padj"

# The bundled table spans many databases, far more terms than read cleanly
# as a ring. Keep the significant ones and take the top 7 per direction;
# which pathways appear is always your choice.
curate <- function(contrast, n = 7) {
  s <- en[en$contrast == contrast & en$padj < 0.05 & is.finite(en$NES), ]
  s <- s[order(s$padj), ]
  rbind(head(s[s$NES > 0, ], n), head(s[s$NES < 0, ], n))
}
en1 <- curate(ctr)
```

The starting point, all defaults:

```{r baseline, fig.alt = "Default volcano-in-ring"}
volcano_ring(da1, en1, title = "YvO", subtitle = ctr)
```

## Colours

Three presets ship: `"default"` (red/blue diverging), `"viridis"`, and
`"okabe"` (colour-blind-safe). Swap one in through `theme`:

```{r palette, fig.alt = "Okabe-Ito palette"}
volcano_ring(da1, en1, theme = volcano_ring_theme(palette = "okabe"))
```

To set your own colours, pass them straight to `volcano_ring_theme()`:
`up`, `down`, `ns` for the points, and `nes_colors` (with optional
`nes_limits`) for the arc ramp.

```{r colours, fig.alt = "Custom point and arc colours"}
my_theme <- volcano_ring_theme(
  up = "#B2182B", down = "#2166AC", ns = "grey80",
  nes_colors = c("#053061", "white", "#67001F"),
  nes_limits = c(-2.5, 2.5)
)
volcano_ring(da1, en1, theme = my_theme)
```

## The volcano

The centre is an ordinary volcano. `p_threshold` and `logfc_threshold`
define the up/down call; `point_size` and `point_alpha` size the cloud;
`label_mode` decides which proteins get text.

```{r volcano, fig.alt = "Bigger points, labelled top hits"}
volcano_ring(da1, en1,
  p_threshold = 0.01, logfc_threshold = log2(1.5),
  point_size = 2.2, point_alpha = 0.6,
  label_mode = "top_per_direction", label_n = 4
)
```

The cloud is scaled to fill the disc inside the ring. When the widest or
tallest points crowd the arcs or the `-log10 p` label, `x_scale` and
`y_scale` (both `1` by default) compress it toward the fold-change axis
without moving the axis annotations:

```{r scale, fig.alt = "Compressed point cloud clearing the ring"}
volcano_ring(da1, en1, x_scale = 0.85, y_scale = 0.9)
```

## The ring

Arc **fill** is NES; arc **thickness** is a magnitude you choose with
`magnitude`, either `"neg_log_padj"` (default) or `"size"` (the gene-set
size). `arc_height_range` sets how short the smallest arc and how tall
the largest arc get, so you can flatten or exaggerate that encoding.

```{r ring, fig.alt = "Size-driven, exaggerated arc heights"}
volcano_ring(da1, en1,
  magnitude = "size",
  arc_height_range = c(0.1, 2.2)
)
```

`arc_order` sets where each arc sits within its half. The up/down split
is always by NES sign; within each half you order by FDR (`"padj"`,
the default) or by enrichment strength (`"nes"`):

```{r order, fig.alt = "Arcs ordered by absolute NES"}
volcano_ring(da1, en1, arc_order = "nes")
```

## Layout

The up/down badges count significant points; turn them off with
`show_counts = FALSE`, or nudge them with `count_x_mult` / `count_y_mult`.
`title`, `subtitle`, and `tag` set the text.

```{r layout, fig.alt = "Counts hidden, custom text"}
volcano_ring(da1, en1,
  show_counts = FALSE,
  title = "Young vs Old", subtitle = "trained", tag = "A"
)
```

## Putting it together in a grid

`volcano_ring_grid()` forwards every one of these arguments to each
panel, so a styled grid is the same knobs passed once:

```{r grid, fig.width = 12, fig.height = 6, fig.alt = "Trained young and old panels"}
da_old <- da[da$contrast == "Training_Old", ]
names(da_old)[names(da_old) == "adj.P.Val"] <- "padj"

g <- volcano_ring_grid(
  list(Young = da1, Old = da_old),
  list(Young = en1, Old = curate("Training_Old")),
  ncol = 2,
  theme = my_theme, arc_order = "nes", show_counts = FALSE
)
g$plot
```

For the column conventions and a limma / DESeq2 / edgeR / fgsea /
clusterProfiler / enrichR cross-walk, see
`vignette("input-contract")`.
