---
title: "Choosing an enrichment database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an enrichment database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(enrichVolcano)
```

## The 21 registered databases

`list_databases()` returns a tibble with one row per registered database.
The full registry ships with the package in
`inst/extdata/db_registry.csv` and is the single source of truth for
the rest of the API.

```{r db-list}
db <- list_databases()
db[, c("name", "display_name", "type", "species", "license")]
```

Twenty-one rows cover six database types. The next section walks
through each type and gives a one-line use-case.

## Database types

### Pathway

Curated biological pathways with explicit gene memberships. Examples:
`hallmark`, `reactome`, `kegg_legacy`, `kegg_medicus`, `wikipathways`,
`biocarta`, `pid`. Use these to answer the question "what biological
process is shifting?". Hallmark is the smallest collection (50 sets)
and gives a fast first pass. Reactome and KEGG_MEDICUS are larger and
more granular; expect more results and more redundancy.

Sub-choices matter. Hallmark sets are deliberately broad (e.g.
HALLMARK_OXIDATIVE_PHOSPHORYLATION has 200 genes). Reactome breaks
the same biology into a hierarchy from "Metabolism" down through
"Respiratory electron transport" to "Complex I biogenesis", with
gene-set sizes from a few thousand to under twenty. KEGG_LEGACY is
the pre-2024 KEGG release; KEGG_MEDICUS is the redesigned 2024
collection with cleaner network logic. WikiPathways skews toward
community-curated metabolism and signalling. BioCarta sets are
small, old, and rarely the right primary source but useful as
cross-reference. PID covers cancer-relevant signalling specifically.

### Ontology

Controlled vocabulary trees from the Gene Ontology and HPO projects.
Examples: `go_bp`, `go_cc`, `go_mf`, `hpo`, and `go_slim` (a bundled
generic-BP slim that trades GO_BP's depth for a compact, readable set).
Use for fine-grained term enrichment when you want to drill into a
specific functional axis.
Expect heavy parent-child redundancy: GO_BP can return 30 enriched
terms that all describe the same process at different depths. Pair
with `ev_collapse()` (see `vignette("pathway-dedup")`) before
plotting.

The three GO branches answer different questions. GO_BP describes
processes ("response to oxidative stress"). GO_CC describes
compartments ("mitochondrial inner membrane"). GO_MF describes
molecular activities ("ATP-binding"). For a proteomics aging study,
GO_BP usually carries the headline biology; GO_CC is the natural
follow-up when localization changes are suspected; GO_MF is best
reserved for biochemistry-heavy studies (enzyme classes, transporter
families). HPO is human-only and links genes to clinical phenotypes;
use it when an aging signature overlaps with a known disease
phenotype set.

### Organelle

Subcellular localization sets. Examples: `mitocarta3`,
`hpa_subcellular`, `uniprot_subcellular`. Use when the question is
"which cellular compartment is affected?". MitoCarta3 has 149 curated
mitochondrial sub-pathways on top of the master mitochondrial gene
list.

MitoCarta3 is the strongest organelle source for mitochondrial work
in human, mouse, and rat. The sub-pathways resolve OXPHOS complexes
individually, separate matrix from intermembrane space, and split
beta-oxidation by chain length. For non-mitochondrial compartments,
pair it with `go_cc` (the GO Cellular Component ontology), `corum`
(see Complex below), and the two whole-cell compartment maps:
`hpa_subcellular` (Human Protein Atlas, human only; 49 imaging-based
"Main location" sets) and `uniprot_subcellular` (top-level UniProt
location compartments parsed from reviewed entries, for human, mouse,
and rat). HPA is antibody-imaging evidence; UniProt is curated from
the literature, so the two complement each other.

### Complex

Curated protein complexes. Example: `corum`. Use when the question is
"which assemblies move together?". CORUM lists experimentally
characterised mammalian complexes, so an enriched set points at a
co-regulated machine (a ribosomal subunit, a respiratory complex)
rather than a loose functional category. It ships as a bundled GMT
for human, mouse, and rat.

### Cell type

Signature sets from sorted cell populations. Examples: `immunesigdb`
(MSigDB C7), `cell_type_sig` (MSigDB C8). Use for tissue-mixing
questions: a bulk muscle biopsy contains satellite cells, immune
infiltrate, and endothelium, and a strong enrichment in one of these
signatures often points at composition shift rather than per-cell
regulation.

### Regulatory

TF and miRNA target sets. Examples: `tft_gtrd`, `mirdb`. Use to ask
"is this driven by a specific transcription factor?". GTRD is built
from ChIP-seq data and gives a binding-based view; miRDB gives the
miRNA seed-match prediction. Treat results as hypothesis-generating.

Pathway databases such as KEGG also carry disease-named entries. The
default `exclude_terms = "DISEASE|CANCER|TUMOR"` regex in
`enrich_volcano()` strips those out so a disease label does not stand
in for a mechanism. To keep them, lower or clear `exclude_terms`.

## A decision flowchart

A short tour for picking a starting point:

* First-pass discovery: `hallmark`. Fifty sets cover most coarse
  biological states and return a readable ring panel.
* Curated canonical pathways: `reactome` or `kegg_medicus`.
* Deep ontology with dedup: `go_bp` plus
  `ev_collapse(scope = "within_db")`.
* Mitochondrial focus: `mitocarta3` plus `go_cc` for cross-check.
* Subcellular localization: `go_cc`, `hpa_subcellular`, or
  `uniprot_subcellular`.
* Protein complexes: `corum`.
* TF activity: `tft_gtrd`.
* Immune phenotype: `immunesigdb`.
* Cell-type composition: `cell_type_sig`.

A typical paper uses two or three sources. Hallmark for the headline
ring, Reactome or GO_BP for finer detail, and one organelle or
regulatory source if the biology calls for it.

Beware of "throw every database at it" mode. Nineteen databases
run against one contrast returns hundreds of overlapping hits and
forces you to triage them post-hoc. Pick two or three sources up
front; the ring panel can only show eight to twelve terms anyway.
For supplementary tables, run everything separately, save the
unfiltered enrichment object, and pick top hits per source.

## Per-database `database_info()` output

`database_info()` prints the registry row for one database in a readable
format. It is the right function to call when a reviewer asks where
a gene set came from.

```{r info-hallmark}
database_info("hallmark")
```

Hallmark is the entry point for almost every analysis in this
package. The sets are coarse enough that a single contrast rarely
hits more than five with strong NES, which keeps the ring panel
readable. Pair with Reactome or GO_BP for finer detail in a supplementary panel.

```{r info-reactome}
database_info("reactome")
```

Reactome ships through `msigdbr` here. It has more than 1600 sets in
the 2024 release. Expect overlap with KEGG, WikiPathways, and BioCarta
on canonical processes; `ev_collapse()` handles that automatically.

```{r info-gobp}
database_info("go_bp")
```

GO_BP is the gold standard for fine-grained biological process
enrichment. It is also the worst offender for parent-child
redundancy. Always pair with Jaccard dedup before plotting.

```{r info-mito}
database_info("mitocarta3")
```

MitoCarta3 ships as a bundled GMT file inside the package, not as a
fetched msigdbr collection. It covers human, mouse, and rat. The 149
MitoPathways subdivide the master mitochondrial gene list into
oxidative phosphorylation complexes, import machinery, beta-oxidation,
and so on. License is CC BY-NC: cite the Rath 2021 paper and stay
non-commercial.

## Multi-species support

The package supports seven species: human, mouse, rat, zebrafish,
fly, yeast, and pig. The species argument flows through both
fetched and bundled databases.

For msigdbr-sourced collections (Hallmark, Reactome, KEGG, GO, etc.),
species names are resolved to scientific names via `msigdbr`'s
ortholog table. Passing `species = "rat"` resolves to
`Rattus norvegicus`. Human gene symbols are the canonical key;
non-human collections are derived by orthology mapping from human
symbols using HCOP and msigdbr's curated tables.

Since msigdbr 10, the underlying database is selected with `db_species`
(`"HS"` or `"MM"`) and the output orthologs with `species`. This package
uses the human database (`db_species = "HS"`) and maps to your `species`,
so for a mouse study you will see a one-time note that mouse results come
from human-derived orthologs rather than the mouse-native MM collections.
That is the intended default; human-keyed sets are the better-curated and
more widely cited reference.

The mapping is one-to-many in principle. One human gene can map to
two rat paralogs; one rat gene can map to none if the orthology is
unresolved. msigdbr handles this by emitting every mapped pair into
the species-specific collection. The practical consequence is that a
rat collection can be slightly larger or smaller than its human
counterpart, and a few rare paralog-only genes will appear in rat
but not human. This is normal and rarely changes top-line
enrichment.

For bundled GMT files (MitoCarta3, CORUM, GO Slim, HPA Subcellular,
UniProt Subcellular), the package selects the species-specific file
from `inst/extdata/gmt/`. The
naming convention is `<db>_<species>.gmt`. If a species file is
missing, `enrich_volcano()` aborts with a clear message rather than
silently falling back to human.

A practical implication: if your gene symbols are not in the
canonical species namespace (rat symbols for a mouse study, for
instance), enrichment will be empty or noise. Re-key the input
before calling `enrich_volcano()`.

## Custom GMT files

For v0.1, pass a named list of pathway lists directly through the
`databases` argument:

```{r custom-db, eval = FALSE}
my_paths <- list(
  CUSTOM_SET_1 = c("ENO1", "PKM", "GAPDH"),
  CUSTOM_SET_2 = c("ATP5F1A", "ATP5F1B", "ATP5F1C")
)
p <- enrich_volcano(
  data       = yvo,
  contrast   = "Aging",
  databases  = list(custom = my_paths)
)
```

The outer list name (here `custom`) becomes the `database` column in
the enrichment output. The inner names become pathway labels in the
ring panel.

To use a GMT file, read it with `fgsea::gmtPathways()` and pass the
resulting list directly, as shown above.

## Licensing

License terms vary by source and apply to redistribution, not to
running the analysis. The registry summarises each:

* **MSigDB Academic**: Hallmark and most C2-C8 collections require
  academic use. Commercial use needs a Broad Institute license.
* **MitoCarta CC BY-NC 4.0**: non-commercial with attribution to
  Rath et al. 2021.
* **GO, Reactome, CORUM, GO Slim, UniProt Subcellular**: open under
  CC BY 4.0. Cite the consortium.
* **HPA Subcellular CC BY-SA 4.0**: open with attribution; the
  share-alike clause applies to redistributed derivatives.
* **KEGG Academic**: free for academic use; commercial use needs
  KEGG Inc. license.
* **miRDB CC BY-NC 4.0**: non-commercial with attribution.

The full license per database is in `list_databases()$license` and
`inst/extdata/db_registry.csv`. The package itself is MIT-licensed;
the licenses on derived gene-set collections are separate.

## Recipes by study type

A few starting points tied to study patterns.

Aging muscle proteomics (YvO 2025 pipeline): Hallmark for the
headline ring, Reactome for finer detail, GO_CC for compartment
shifts, MitoCarta3 for OXPHOS detail. Four sources, run separately,
combined with `scope = "within_db"` so each source contributes a
distinct slice to the ring.

Cancer vs healthy serum proteomics (CvH 2026 pipeline): Hallmark and
Reactome as primary; `go_cc` and `corum` to separate secreted from
cell-leak signals; ImmuneSigDB to flag immune-driven changes.

High vs low responder training studies (HRvLR 2026 pipeline):
Hallmark and Reactome with `scope = "cross_db"` since responders and
non-responders share most biology and the goal is non-redundant
top-line hits. Add `tft_gtrd` to ask whether a TF program separates
the two groups.

H9c2 mitochondrial transplant (Mito 2026 conceptual pipeline):
MitoCarta3 as the primary source; GO_CC for membrane vs matrix
splits; Reactome for OXPHOS and TCA cycle detail. Hallmark is too
coarse for the mitochondrial focus and best dropped.

These recipes are starting points, not prescriptions. The right
database mix depends on the biology and the audience. For a methods
section, name every source and cite the registry version.

## Comparing across runs

Re-running `enrich_volcano()` with the same data and the same
database set returns the same enrichment table. `msigdbr` ships
versioned releases, so locking the package version locks the gene
sets. For long-running projects, record the `msigdbr` version
alongside the analysis script:

```{r record-version, eval = FALSE}
sessionInfo()$otherPkgs$msigdbr$Version
```

Bundled GMT files are versioned through the package version. The
2024 release of MitoCarta3 ships with v0.1 of enrichVolcano. A
future Mitocarta3.1 will bump the enrichVolcano minor version.

Next: how to style your plots — see `vignette("customising")`.

## Compartment enrichment

Subcellular-compartment enrichment is an established interpretive layer for
proteomics differential abundance (Subramanian et al. 2005; Thul et al. 2017;
Binder et al. 2014). The `compartments` keyword enriches three reproducible,
versioned sources at once:

```{r compartments, eval = FALSE}
p <- enrich_volcano(yvo, contrast = "Aging", databases = "compartments")
# expands to c("go_cc", "corum", "mitocarta3"), each species-filtered
```

`go_cc` (GO Cellular Component, msigdbr-versioned), `corum` (protein
complexes), and `mitocarta3` (mitochondrial sub-compartments) are curated and
version-stable, so results reproduce across runs. Three caveats from the
literature, all handled by the package defaults: annotation and abundance bias
is controlled by using the quantified proteome as the ORA background;
multi-localization redundancy is handled by `ev_collapse()`; and GO:CC
parent-child overlap is reduced by the size filter and dedup
(Reimand et al. 2019).
