---
title: "enrichVolcano"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{enrichVolcano}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  dpi = 72
)
```

enrichVolcano draws figures from enrichment results you have already
computed. Every workflow has the same three steps:

1. `as_enrichment()` converts results from any supported tool into one
   validated `enrichment` object.
2. `dedup()` (optional) flags redundant terms so figures show each signal
   once.
3. A plot function draws from the object: `volcano_ring()` puts a
   differential-abundance volcano inside a ring of enrichment terms, and
   `nes_scatter()` compares two contrasts term by term.

## Install

```{r install, eval = FALSE}
install.packages("remotes")
remotes::install_github("Dustyn-T-Lewis/enrichVolcano")
```

## Quick start

The package ships one example study: fgsea results for four contrasts across
five gene-set collections, and the matching limma differential-abundance
table. The collections are MSigDB Hallmark, KEGG MEDICUS, Reactome and GO:BP
from msigdbr 26.1.0, and the GO Consortium's generic GO slim (release
2026-07-26).

```{r read}
library(enrichVolcano)

fgsea_path <- system.file("extdata", "examples", "yvo_fgsea.csv.gz", package = "enrichVolcano")
da_path <- system.file("extdata", "examples", "yvo_da.csv.gz", package = "enrichVolcano")

ex <- as_enrichment(read.csv(fgsea_path))
ex
```

`as_enrichment()` recognised fgsea output from its columns and split the
table by its `contrast` column. The DA table needs one contrast and a `padj`
column:

```{r da}
da <- read.csv(da_path)
names(da)[names(da) == "adj.P.Val"] <- "padj"
young <- da[da$contrast == "Training_Young", ]
```

```{r ring, fig.alt = "Volcano of Training_Young proteins inside a ring of enriched Hallmark and GO slim terms"}
volcano_ring(young, ex, contrast = "Training_Young", title = "Training, young")
```

Arc fill is NES and arc height is \(-\log_{10}\)(padj). Up-regulated terms sit
on the upper half, down-regulated ones on the lower. Ticks run from each arc
to its leading-edge proteins in the volcano. By default the ring takes the
twelve most significant terms (padj < 0.05), in either direction, from
Hallmark and GO slim;
`databases`, `term_threshold`, `n_terms` and `terms` change the selection.

## Convert your results

Every tool below ends in the same object. Pass one table per contrast as a
named list, or one table with a `contrast` column.

### fgsea

```{r fgsea, eval = FALSE}
res <- fgsea::fgsea(pathways, stats = ranks)
x <- as_enrichment(list(Aging = res), database = "Hallmark")
```

The score is the normalised enrichment score (NES): how far the gene set sits
toward the top or bottom of the ranking, scaled so sets of different sizes
compare. `database` labels every row when the table has no `database` column.

### clusterProfiler

```{r gsea-result, eval = FALSE}
res <- clusterProfiler::gseGO(gene_list, ont = "BP", OrgDb = org.Hs.eg.db::org.Hs.eg.db)
x <- as_enrichment(list(Aging = res), database = "GO:BP")
```

The `gseaResult` object is read directly. The score is NES and the leading
edge comes from `core_enrichment`.

### limma: fry and mroast

```{r fry-run, eval = FALSE}
res <- limma::fry(y, index = gene_sets, design = design, contrast = contrast)
x <- as_enrichment(list(Aging = res), database = "Hallmark")
```

fry and mroast are **self-contained** tests: they ask whether the genes in a
set changed at all, without reference to genes outside it. Neither reports an
effect size like NES, or a leading edge. The score is therefore
\(-\log_{10}\)(FDR), signed by the reported direction, and the figure's legend
says so. The table below has fry's shape; the numbers are invented to show the
conversion.

```{r fry}
fry_res <- data.frame(
  NGenes = c(200, 150, 180, 160),
  Direction = c("Up", "Up", "Down", "Down"),
  PValue = c(1e-6, 4e-4, 2e-5, 3e-3),
  FDR = c(4e-6, 8e-4, 4e-5, 4e-3),
  PValue.Mixed = c(1e-7, 1e-4, 1e-5, 1e-3),
  FDR.Mixed = c(4e-7, 2e-4, 2e-5, 1e-3),
  row.names = c(
    "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_MYOGENESIS",
    "HALLMARK_INFLAMMATORY_RESPONSE", "HALLMARK_APOPTOSIS"
  )
)
fry_x <- as_enrichment(list(Training_Young = fry_res), database = "Hallmark")
```

```{r fry-ring, fig.alt = "Volcano ring drawn from fry results, with arcs coloured by signed -log10 FDR and no tick lines"}
volcano_ring(young, fry_x, title = "fry")
```

With no NES to encode, arc fill carries the signed \(-\log_{10}\)(FDR) and arc
height switches to set size, so the two channels show different facts.

### limma: camera and cameraPR

```{r camera, eval = FALSE}
res <- limma::camera(y, index = gene_sets, design = design, contrast = contrast)
x <- as_enrichment(list(Aging = res), enrichment_test = "camera", database = "Hallmark")
```

camera and cameraPR are **competitive** tests, like fgsea: they ask whether a
set changed more than the genes outside it. Their default output has
identical columns, so `as_enrichment()` cannot tell them apart and asks you to
name the test with `enrichment_test`. camera run with
`inter.gene.cor = NA` adds a `Correlation` column and is recognised
automatically.

### Over-representation analysis

```{r ora, eval = FALSE}
up <- enrichR::enrichr(up_genes, "GO_Biological_Process_2023")[[1]]
down <- enrichR::enrichr(down_genes, "GO_Biological_Process_2023")[[1]]
up$direction <- "up"
down$direction <- "down"
x <- as_enrichment(
  list(Aging = rbind(up, down)),
  enrichment_test = "ora",
  term = "Term", padj = "Adjusted.P.value", p = "P.value", leading_edge = "Genes"
)
```

Over-representation asks whether a gene list overlaps a set more than chance
predicts; it has no direction of its own. Run it on up- and down-regulated
genes separately and stack the two with a `direction` column. The score is
signed \(-\log_{10}\)(padj).

### Anything else

```{r custom, eval = FALSE}
x <- as_enrichment(
  list(Aging = my_table),
  enrichment_test = "custom",
  term = "set", score = "stat", padj = "q", size = "n", score_type = "z"
)
```

Name the columns and, optionally, what the score is; `score_type` becomes the
legend title.

## Deduplicate

Related gene sets often reach significance together. In GO:BP or Reactome,
one biological signal can light up a dozen overlapping terms. `dedup()` flags
all but one per cluster as `"redundant"`, and plots hide redundant terms by
default (`collapse = TRUE`). No row is removed and no p-value changes: the
multiple-testing correction already covered every term, so redundancy is a
display decision.

```{r dedup, eval = FALSE}
gene_sets <- split(msig$gene_symbol, msig$gs_name)
x <- dedup(x, gene_sets)
x <- dedup(x, gene_sets, similarity = "jaccard")
x <- dedup(x, gene_sets, method = "collapse_pathways", stats = list(Aging = ranks))
```

Within each contrast and database, significant terms are walked from the
smallest padj. A term is redundant when its gene set is similar enough to a
term already kept:

- `similarity = "combined"` (default, cutoff 0.375) is the EnrichmentMap
  coefficient: the mean of the Jaccard index and the overlap coefficient. The
  overlap coefficient catches a small set nested inside a large one, which
  Jaccard alone misses.
- `similarity = "jaccard"` (cutoff 0.5) is shared genes over all genes in
  either set.

`method = "collapse_pathways"` is a different kind of rule. It calls
`fgsea::collapsePathways()`, which asks whether a term is still enriched once
the genes of a more significant term are conditioned on. That is a new
statistical test, so it needs the ranking each contrast was tested on, works
only for ranked GSEA results, and should be reported as its own analysis.
Call `set.seed()` first; it permutes.

The example object carries flags computed upstream with the Jaccard rule. They
pass straight through `as_enrichment()`:

```{r flags}
table(ex@results$dedup_status, ex@results$database, useNA = "ifany")
```

Hallmark and GO slim lose nothing because both are built to be non-redundant.
The flags matter once you draw from GO:BP or Reactome:

```{r reactome, fig.alt = "Volcano ring drawn from Reactome terms with redundant terms hidden"}
volcano_ring(young, ex, contrast = "Training_Young", databases = "Reactome", title = "Reactome")
```

## Compare two contrasts

`nes_scatter()` plots every term's score in one contrast against its score in
another. It answers two kinds of question, set by `comparison`.

The default, `comparison = "concordance"`, asks whether two conditions
regulate the same pathways. Terms in the shaded top-right and bottom-left
quadrants move the same way in both.

```{r concordance, fig.width = 6.5, fig.height = 7.5, fig.alt = "Scatter of NES in trained young against trained old, with most significant terms in the concordant quadrants"}
nes_scatter(ex, "Training_Young", "Training_Old")
```

`comparison = "reversal"` asks whether one condition undoes another. The
unshaded quadrants hold the answer: a pathway raised by aging and lowered by
training, or the reverse. The reference line becomes \(y = -x\).

```{r reversal, fig.width = 6.5, fig.height = 7.5, fig.alt = "Scatter of NES in aging against trained old, with most significant terms in the reversed quadrants"}
nes_scatter(ex, "Aging", "Training_Old", comparison = "reversal")
```

A term counts as significant in a contrast when its padj is below
`p_threshold` (0.05); the colour says whether that holds in one contrast or
both. Pathways are judged on the false discovery rate alone. Scores that
combine effect size with p, such as the \(\pi\)-value, are defined for single
proteins, not gene sets. The corner counts and the headline share use
significant terms only; Spearman's \(\rho\) uses every plotted term, with a 95%
interval when the correlation package is installed.

`color_by` and `shape_by` take any per-term column, for example
`color_by = "database"`, and `databases`, `collapse` and `label_min_size`
work as they do for the ring.

## Change the look

`volcano_ring_theme()` holds the palette; the plot arguments hold the layout.

```{r palette, eval = FALSE}
volcano_ring(young, ex,
  contrast = "Training_Young",
  theme = volcano_ring_theme(palette = "okabe"),
  label_mode = "top_per_direction", label_n = 4
)
```

The arguments you will reach for most:

- `magnitude`: what arc height encodes, `"neg_log_padj"` or `"size"`.
- `arc_height_range`: shortest and tallest arc.
- `arc_order`: order arcs by `"padj"` or by absolute score within each half.
- `x_scale`, `y_scale`: compress the volcano when points crowd the ring.
- `label_mode`, `label_n`, `label_genes`: which proteins get text.
- `show_counts`, `disc_color`: count badges and a tinted central disc.

`volcano_ring_grid()` draws one ring per contrast and collects a shared
legend; extra arguments pass to every ring.

```{r grid, fig.width = 10, fig.height = 5.5, fig.alt = "Two volcano rings side by side for trained young and trained old"}
g <- volcano_ring_grid(da, ex, contrasts = c("Training_Young", "Training_Old"), ncol = 2)
g$plot
```
