---
title: "Introduction to enrichVolcano"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to enrichVolcano}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What this package does

`enrichVolcano` builds one figure with two panels: a volcano plot of
differential abundance results and a ring plot of pathway enrichment
for the same contrast. Input can come from `proteoDA`, `DEP`, `limma`,
`DESeq2`, `edgeR`, `MSstats`, `proDA`, `DEqMS`, MaxQuant, or Perseus;
`ev_validate()` auto-detects the format and returns a tidy long
tibble. The hero function `enrich_volcano()` returns a `patchwork`
object that you can save with `ggplot2::ggsave()` or further compose
with other plots.

```{r setup}
library(enrichVolcano)
```

## A 30-second example

Load the bundled YvO (young vs. old skeletal muscle) result set. It
ships with the package after install; during development you can read
it from the test fixtures.

```{r load-data}
fixture <- system.file(
  "extdata", "examples", "yvo_tidy.rds",
  package = "enrichVolcano"
)
if (!nzchar(fixture)) {
  fixture <- "../tests/testthat/fixtures/yvo_tidy.rds"
}
yvo <- readRDS(fixture)
yvo <- yvo[!is.na(yvo$gene) & !is.na(yvo$contrast), ]
head(yvo, 3)
unique(yvo$contrast)
```

The fixture has 400 protein rows across four contrasts (Aging,
Training_Young, Training_Old, Interaction). Each row already has
`logFC`, `P.Value`, and `adj.P.Val`.

The default registered databases (`hallmark`, `reactome`, `go_bp`)
fetch from `msigdbr`, which needs an internet connection. For a
self-contained example we build a small mock pathway list from the
genes in the fixture.

```{r mock-db}
set.seed(42)
genes <- unique(yvo$gene)
mock_msigdb <- list(
  HALLMARK_GLYCOLYSIS         = sample(genes, 30),
  HALLMARK_OXIDATIVE_PHOS     = sample(genes, 25),
  HALLMARK_APOPTOSIS          = sample(genes, 20),
  HALLMARK_MYOGENESIS         = sample(genes, 35),
  REACTOME_CITRIC_ACID_CYCLE  = sample(genes, 15)
)
```

Now call `enrich_volcano()` with one contrast and the mock database:

```{r hero-call, fig.alt = "Volcano panel above an enrichment ring panel for the Aging contrast."}
fig <- enrich_volcano(
  data       = yvo,
  contrast   = "Aging",
  databases  = list(mock = mock_msigdb),
  enrich_padj = 1,
  enrich_mode = "fgsea",
  ring   = list(max_terms = 5)
)
fig
```

`enrich_padj = 1` keeps every pathway in the ring for the demo; in
practice you would use the default 0.05. The `mock_msigdb` here stands
in for the real Hallmark or Reactome collections.

## UniProt-anchored input

Point `ev_read_contrasts()` at a folder of per-contrast CSVs (each with a
UniProt accession column, `logFC`, and a p-value). The file name becomes the
contrast label; accessions are mapped to gene symbols for enrichment, and the
mapping summary is available via `ev_idmap_report()`.

```{r uniprot-input, eval = FALSE}
res <- ev_read_contrasts("path/to/contrast_csvs", species = "human")
ev_idmap_report(res)
enrich_volcano(res, contrast = "Cancer_vs_Healthy",
               label_mode = "top_per_direction", label_n = 8)
```

Volcano labels are off by default; choose `label_mode` =
`"top_per_direction"`, `"top_total"`, `"all_significant"`, or `"explicit"`,
and rank candidates by `label_rank_by` = `"significance"` or `"logfc"`.

## Anatomy of the output

The volcano sits at the centre: `logFC` on the x-axis, `-log10(pi_eq2)`
on the y-axis. A point is called up or down when its pi-score clears
`p_threshold` (default 0.05) and its `logFC` is positive or negative;
the panel reports the up and down counts in coloured boxes rather than
labelling individual genes. (For a labelled standalone volcano with
`ggrepel`, call `ev_volcano()` — see `vignette("customising")`.)

The enrichment ring wraps the volcano, drawn with
`ggforce::geom_arc_bar` in a fixed Cartesian layout (no `coord_polar`).
Up-regulated pathways cluster on the right (centred at 90°) and
down-regulated on the left (270°), so direction reads at a glance. Each
arc's outer radius encodes `-log10(padj)` (taller = more significant)
and its fill encodes NES on a colourblind-safe blue-white-red diverging
scale. Short radial ticks under each arc mark its leading-edge genes,
coloured by the sign of their fold change.

The contrast name is the title. The composite is a `patchwork` object,
so it supports the `+`, `/`, and `|` operators for further layout.

Two attributes hold the intermediate state:

```{r inspect-attrs}
str(attr(fig, "ev_data"), max.level = 1)
attr(fig, "ev_call")
```

`ev_data` is a list with four tibbles: the validated input, the
pi-score table, the raw `ev_enrich()` output, and the post-dedup
enrichment table. `ev_call` is the literal `match.call()`, so a
reviewer can re-run the exact figure from the object alone.

## The pipeline under the hood

`enrich_volcano()` chains the exported pipeline stages and then draws
the canonical composite:

1. `ev_validate(data)` detects the input format and renames columns
   to `gene`, `contrast`, `logFC`, `P.Value`, `adj.P.Val`.
2. `pi_score(data, variant = "eq2")` computes the Xiao 2014
   pi-score and writes it to `pi_eq2`.
3. `adjust_p(data, method = "BH")` recomputes `adj.P.Val` from the
   raw p-values within each contrast.
4. `ev_enrich(data, contrast, databases)` runs `fgsea` (and optionally
   `fora`) against the requested pathway sets, ranking genes by the
   signed statistic `sign(logFC) * -log10(P)`.
5. `ev_collapse(enrich, method = "jaccard")` removes redundant
   pathways with overlapping leading-edge gene sets.
6. `ev_volcano_ring(volcano_sub, enrich_sub)` draws one volcano-in-ring
   composite per contrast; `patchwork::wrap_plots()` arranges multiple
   contrasts and attaches `ev_data` and `ev_call`.

Stages 1-5 are exported, so a real analysis can stop between them — a
common pattern is to compute pi-scores once, write the table to disk,
and re-enter at `ev_enrich()` with different database choices. For a
panel layout you assemble yourself, the package also ships the
standalone trio `ev_volcano()` + `ring_plot()` + `ev_compose()`, which
expose the tunable label and ring-encoding options the opinionated
composite keeps fixed (see `vignette("customising")`).

## What next

* `vignette("scoring")` covers the pi-score variants and the four
  p-value adjustment methods.
* `vignette("pathway-dedup")` explains Jaccard collapsing, scope
  (within vs. across databases), and when to use
  `fgsea::collapsePathways` instead.
* `vignette("databases")` lists the registered gene-set sources and
  shows how to add a custom GMT.
* `vignette("customising")` walks through theme overrides, palette
  swaps, and per-panel ggplot edits.
* `vignette("faq")` answers questions about MaxQuant ratios,
  contrast naming, and large figure layouts.
