---
title: "FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(enrichVolcano)
```

## Why is my ring plot empty?

Four likely causes.

The classic one is the **wrong ranking statistic**. GSEA needs a
*signed* score so up- and down-regulated genes sit at opposite ends of
the ranked list (Subramanian et al. 2005; Reimand et al. 2019). Ranking
by an unsigned pi-score collapses everything to one tail and yields
near-empty rings. The package defaults to `rank_by = "signed_p"`
(`sign(logFC) * -log10(P)`) and now refuses an unsigned pi-score with an
`ev_unsigned_ranking` error rather than silently mis-ranking. If you see
that error, drop the `rank_by` override and use the default. See
`vignette("scoring")` for the full explanation.

Second, no pathways passed the `enrich_padj` filter. The default is
0.05, which is strict on a small contrast or a small fixture. Inspect
the padj distribution on the figure object:

```{r empty-ring, eval = FALSE}
p <- enrich_volcano(data, contrast = "ctr")
summary(attr(p, "ev_data")$enrichment$padj)
```

If the smallest padj is above 0.05, raise the cutoff:

```{r empty-relax, eval = FALSE}
p <- enrich_volcano(data, contrast = "ctr", enrich_padj = 0.25)
```

Third, your gene symbols do not match the database species. msigdbr
collections key on human symbols for `species = "human"`, mouse
symbols for `"mouse"`, and so on. If your input has rat symbols but
you passed `species = "human"`, the overlap is empty by construction.

Fourth, all pathways were dropped by `ev_collapse()`. Loosen the
dedup cutoff by passing a higher value through the `dedup` list:

```{r empty-collapse, eval = FALSE}
p <- enrich_volcano(
  data, contrast = "ctr",
  dedup = list(method = "jaccard", cutoff = 0.9, scope = "within_db")
)
```

A cutoff close to 1 keeps almost everything. If the ring still
fills, the original cutoff was eating real biology; if it stays
empty, the issue is upstream of dedup.

## Can I use DESeq2 / edgeR / Perseus output directly?

Yes. `ev_validate()` auto-detects input formats from column
patterns. Supported sources: tidy long, wide, DEP `data.frame`,
proteoDA result tables, limma `topTable`, DESeq2 `results`, edgeR
`topTags`, MSstats `comparisonResult`, proDA `test_diff`, DEqMS,
MaxQuant ratio outputs, and Perseus exported tables. Pass the data
frame directly to `enrich_volcano()`.

Perseus reports its p-value as `-Log10 p-value`. `ev_validate()`
back-transforms with a `cli_inform` notice so you see what happened.
MaxQuant ratio outputs lack an FDR column; BH is synthesized at
validation time, also with a notice.

DEP `SummarizedExperiment` objects and proteoDA `DAList` objects are
not yet directly supported in v0.1. Extract `rowData()` (DEP) or
`$results` (proteoDA) to a data frame first:

```{r dep-extract, eval = FALSE}
df <- as.data.frame(SummarizedExperiment::rowData(dep_se))
p  <- enrich_volcano(df, contrast = "ctr")
```

## How do I force-label specific genes on the volcano?

`enrich_volcano()` now forwards `label_mode`, `label_n`, `label_rank_by`, and
`label_genes` to the composite volcano, so you can label directly from the
hero call. Use `label_mode = "explicit"` to pin specific genes regardless of
rank:

```{r force-label, eval = FALSE}
p_volcano <- ev_volcano(
  data, contrast = "ctr",
  label_mode  = "explicit",
  label_genes = c("GENE1", "GENE2")
)
```

Then build the ring and compose:

```{r force-compose, eval = FALSE}
enr <- ev_enrich(data, contrast = "ctr", databases = "hallmark")
p_ring <- ring_plot(enr, contrast = "ctr")
p <- ev_compose(list(ctr = p_volcano), list(ctr = p_ring))
```

## Why don't I see KEGG pathways even though I requested them?

Four checks.

* `msigdbr` is installed and reasonably current. KEGG_MEDICUS was
  added to MSigDB in 2024; older `msigdbr` releases ship
  `kegg_legacy` only.
* The species matches. KEGG via msigdbr supports the seven species
  in the registry. Pass `species = "human"` (or the matching value)
  alongside `databases = "kegg_medicus"`.
* `database_info("kegg_medicus")` confirms the database is registered
  and shows the source. If the source reads `msigdbr`, the lookup
  goes through `msigdbr::msigdbr()`.
* The default `exclude_terms = "DISEASE|CANCER|TUMOR"` regex inside
  `ev_enrich()` strips disease-named KEGG entries out of pathway
  databases. KEGG includes hundreds of disease-named entries; the
  default filters them so a disease label does not stand in for a
  mechanism. To keep them, call `ev_enrich()` directly:

```{r kegg-disease, eval = FALSE}
enr <- ev_enrich(
  data, contrast = "ctr",
  databases     = "kegg_medicus",
  exclude_terms = NULL
)
p_ring <- ring_plot(enr, contrast = "ctr")
```

## How do I add my own gene set file?

For v0.1, pass a named list of named lists. The outer name is the
database name; the inner names are pathway names; the values are
character vectors of gene symbols.

```{r custom-sets, eval = FALSE}
my_paths <- list(
  CUSTOM_SET_1 = c("GENE1", "GENE2", "GENE3"),
  CUSTOM_SET_2 = c("GENE4", "GENE5")
)
p <- enrich_volcano(
  data      = data,
  contrast  = "ctr",
  databases = list(custom = my_paths)
)
```

To load from a GMT file, read it with `fgsea::gmtPathways()` and
pass the result through the same shape:

```{r gmt-load, eval = FALSE}
my_paths <- fgsea::gmtPathways("my_sets.gmt")
p <- enrich_volcano(
  data      = data,
  contrast  = "ctr",
  databases = list(custom = my_paths)
)
```

Reading the file yourself and passing the list keeps the gene-set
source explicit and under your control.

## My data is rat (or mouse / fly / zebrafish / yeast / pig). What do I pass?

Set `species` and the registered databases follow:

```{r species, eval = FALSE}
p <- enrich_volcano(
  data      = data,
  contrast  = "ctr",
  species   = "rat",
  databases = "mitocarta3"
)
```

Supported species strings: `"human"`, `"mouse"`, `"rat"`,
`"zebrafish"`, `"fly"`, `"yeast"`, `"pig"`. msigdbr handles
ortholog mapping for fetched collections; bundled GMTs (MitoCarta3,
CORUM, GO Slim, HPA Subcellular, UniProt Subcellular) ship per-species
files where the species is supported.

Input gene symbols must match the species namespace. Rat symbols for
a rat study, mouse symbols for a mouse study. Mixing namespaces
silently returns empty enrichment.

## Why is the rendered PDF different from the screen preview?

Three sizing rules to remember.

`ggsave()` defaults to inches. Pass `units = "mm"` (or `"cm"`)
explicitly when sizing for a journal column width specified in
millimetres.

Patchwork inherits panel sizes from its components. If the volcano
and ring panels carry very different intrinsic aspect ratios, the
on-screen preview at one device size will not match the saved file at
a different size. Set fixed sizes with `&` and `theme(aspect.ratio =
1)` if the layout matters.

The ring uses `coord_fixed()`. Its 1:1 aspect ratio is locked at
draw time and ignores save dimensions. The ring will always be a
circle; the patchwork resizes the volcano to fill the remaining
space.

For predictable output, save to PDF first at the exact width and
height you need, then check the PDF in a viewer rather than the
RStudio plots pane.

## How do I cite this package?

```{r cite, eval = FALSE}
citation("enrichVolcano")
```

A full citation needs four references: this package, Xiao et al.
2014 for the pi-score, Korotkevich et al. 2021 for fgsea, and
Benjamini and Hochberg 1995 for BH adjustment. All four are in
`inst/REFERENCES.bib` and ship with the package install.

If you used a specific database, cite its primary source as well.
`database_info(name)$citation_key` resolves to the BibTeX key in
`REFERENCES.bib`.
