---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  dpi = 96
)
```

`enrichVolcano` draws one composite figure from two tidy tables that you
have already computed: a differential-abundance (DA) table and an
enrichment table. Computing the enrichment is delegated to whichever
upstream tool fits your study — `fgsea`, `clusterProfiler`, `enrichR`, or
your own.

This vignette walks the end-to-end path on the bundled YvO proteomics
example (Young vs Old human plasma; one contrast, `Training_Young`).

## Install

If you do not have remotes installed:

```{r install-remotes, eval = FALSE}
install.packages("remotes")
```

Then:

```{r install, eval = FALSE}
remotes::install_github("Dustyn-T-Lewis/enrichVolcano")
```

## Read the two tables

The package ships small slices of four datasets under
`inst/extdata/examples/`. `system.file()` resolves to the install path.

```{r read}
library(enrichVolcano)

da_path <- system.file("extdata", "examples", "yvo_da.csv",
  package = "enrichVolcano"
)
en_path <- system.file("extdata", "examples", "yvo_enrichment.csv",
  package = "enrichVolcano"
)

da <- read.csv(da_path)
en <- read.csv(en_path)

str(da, max.level = 1, give.attr = FALSE)
str(en, max.level = 1, give.attr = FALSE)
```

The YvO DA table is a limma moderated-t output: `gene`, `contrast`,
`logFC`, `P.Value`, `adj.P.Val`, `pi_eq2`. The enrichment table is in the
fgsea schema with a per-row contrast column: `contrast`, `database`,
`pathway`, `NES`, `padj`, `size`. There is one contrast here
(`Training_Young`).

## Slice to a single contrast

`volcano_ring()` draws one composite per call. Multi-contrast composing
is what `volcano_ring_grid()` is for.

```{r slice}
ctr <- "Training_Young"
da1 <- da[da$contrast == ctr, , drop = FALSE]
en1 <- en[en$contrast == ctr, , drop = FALSE]
nrow(da1)
nrow(en1)
```

The enrichment slice spans many databases, far more terms than read
cleanly as a ring. Curation is yours to do: keep the significant terms
and take the top few per direction.

```{r curate}
sig <- en1[en1$padj < 0.05 & is.finite(en1$NES), , drop = FALSE]
sig <- sig[order(sig$padj), ]
en1 <- rbind(head(sig[sig$NES > 0, ], 7), head(sig[sig$NES < 0, ], 7))
nrow(en1)
```

## Map column names

The YvO DA carries `adj.P.Val`, not `padj`. The shared `padj_col`
argument names *both* the adjusted-p column on the DA side and the
adjusted-p column on the enrichment side, so they need to agree. Two
options: pass `padj_col = "adj.P.Val"` (but then the enrichment side
needs that column too), or rename the DA column to `padj` once at the
boundary. The second is simpler.

```{r rename}
names(da1)[names(da1) == "adj.P.Val"] <- "padj"
```

## Draw the composite

```{r ring, fig.alt = "Volcano in ring for YvO Training_Young"}
suppressMessages(
  volcano_ring(da1, en1, title = "YvO", subtitle = ctr)
)
```

The figure puts the volcano in the centre. NES-coloured arcs around it
encode each enriched pathway: arc fill is NES (red up, blue down by
default), arc thickness is `-log10(padj)`, and the up/down split lives on
the right vs left semicircle. The pre-filtering of which pathways appear
is up to you — every row of `enrich_df` becomes one arc.

## Tick lines to the leading edge

A pathway earns its place because a handful of proteins drive it. When the
enrichment table carries those leading-edge genes, `volcano_ring()` draws a
thin tick from each arc to its genes inside the volcano, so a pathway reads
back to the points that produced it. The YvO enrichment table above has no
gene column, so the arcs above draw without ticks.

Here the two example arcs carry the strongest up- and down-regulated
proteins from `da1` as their leading edge, and the ticks appear:

```{r ticks, fig.alt = "Volcano in ring with leading-edge tick lines"}
strongest <- function(direction, n = 6) {
  hits <- da1[da1$padj < 0.05 & sign(da1$logFC) == direction, ]
  head(hits[order(hits$padj), "gene"], n)
}
en_ticks <- data.frame(
  pathway = c("Up example set", "Down example set"),
  NES = c(2.1, -1.9),
  padj = c(0.002, 0.006),
  size = c(6L, 6L),
  leading_edge = c(
    paste(strongest(1), collapse = ";"),
    paste(strongest(-1), collapse = ";")
  )
)
suppressMessages(volcano_ring(da1, en_ticks, title = "Leading-edge ticks"))
```

`genes_col` is auto-detected from `leading_edge`, `leadingEdge`,
`core_enrichment`, or `Genes`; `vignette("input-contract")` lists the
conventions each upstream tool uses.

## A multi-contrast grid

When you have several contrasts to show side by side, pass two named
lists to `volcano_ring_grid()`. The function calls `volcano_ring()` once
per panel and composes them with `patchwork::wrap_plots()`, collecting one
shared NES legend. Here are the trained young and old contrasts, each
curated to its top pathways:

```{r grid, fig.width = 12, fig.height = 7, fig.alt = "Trained young and old panels"}
curate <- function(contrast, n = 7) {
  s <- en[en$contrast == contrast & en$padj < 0.05 & is.finite(en$NES), ]
  s <- s[order(s$padj), ]
  rbind(head(s[s$NES > 0, ], n), head(s[s$NES < 0, ], n))
}
da_old <- da[da$contrast == "Training_Old", , drop = FALSE]
names(da_old)[names(da_old) == "adj.P.Val"] <- "padj"

g <- suppressMessages(volcano_ring_grid(
  list(Young = da1, Old = da_old),
  list(Young = en1, Old = curate("Training_Old")),
  ncol = 2
))
g$plot
```

## Common knobs

The full argument list is in `?volcano_ring`, but four come up
constantly:

- `magnitude = "neg_log_padj"` (default) or `"size"` sets what controls
  arc thickness.
- `genes_col` and `genes_sep` override the tick-line auto-detect.
  Auto-detect walks `leading_edge`, `leadingEdge`, `core_enrichment`,
  `Genes` in that order.
- `label_mode` (`"none"`, `"top_per_direction"`, `"by_significance"`,
  `"by_genes"`) controls which volcano points are labelled.
- `theme = volcano_ring_theme(palette = "viridis")` swaps the palette
  without re-passing every aesthetic.

`vignette("customizing")` walks the colours, volcano, ring, and layout
knobs section by section.

## What this package does not do

`enrichVolcano` does not run the enrichment for you. If you need to
compute it, here are the canonical paths:

```r
# fgsea (Korotkevich 2021, DOI 10.1101/060012):
ranks <- df$t; names(ranks) <- df$gene
res <- fgsea::fgseaMultilevel(pathways, stats = ranks)

# clusterProfiler::gseGO (Yu 2012; Wu 2021):
res <- clusterProfiler::gseGO(
  geneList = sort(setNames(df$logFC, df$gene), decreasing = TRUE),
  ont = "BP", OrgDb = org.Hs.eg.db::org.Hs.eg.db
)

# enrichR (Chen 2013; Kuleshov 2016):
res <- enrichR::enrichr(genes, "GO_Biological_Process_2023")[[1]]
```

The result of any of these is a tidy frame you can pass straight to
`volcano_ring()`. See `vignette("input-contract")` for column names that
work out of the box.
