---
title: "Input contract"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input contract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`volcano_ring()` takes two tidy frames: a differential-abundance (DA)
frame and an enrichment frame. Every column the function consumes is
named via an argument with a sensible default, so the path of least
resistance is to use limma + fgsea conventional names. Everything else
is a rename at the call site.

```{r overview, eval = FALSE}
library(enrichVolcano)

volcano_ring(
  volc_df, enrich_df,

  # volcano frame
  gene_col = "gene", # protein/gene identifier
  logfc_col = "logFC", # log fold change
  pval_col = "P.Value", # raw p
  padj_col = "padj", # adjusted p (also used for enrich_df)

  # enrichment frame
  term_col = "pathway", # pathway / term name
  nes_col = "NES", # signed effect size for the term
  size_col = "size", # only required when magnitude = "size"
  genes_col = NULL, # NULL triggers tick auto-detect (Q4)
  genes_sep = NULL, # NULL defers to the auto-detected default

  # thresholds + encoding
  p_threshold = 0.05,
  logfc_threshold = 0,
  magnitude = c("neg_log_padj", "size")
)
```

## The volcano frame

`volcano_ring()` needs `gene`, `logFC`, and a p-value column. If a
`padj` column is present it gates the up/down call; if not, the raw
`pval` is used.

| Column | Type | Required? | Notes |
|---|---|---|---|
| `gene_col` | character | yes | one row per gene per contrast |
| `logfc_col` | numeric | yes | log-scale effect |
| `pval_col` | numeric in [0, 1] | yes | raw or moderated |
| `padj_col` | numeric in [0, 1] | optional | when present, gates significance |

### Cross-walk

| Upstream | `gene_col` | `logfc_col` | `pval_col` | `padj_col` | Note |
|---|---|---|---|---|---|
| limma `topTable()` | `Gene` or row name | `logFC` | `P.Value` | `adj.P.Val` | rename `adj.P.Val` to `padj`, or pass `padj_col = "adj.P.Val"` |
| DESeq2 `results()` | row name | `log2FoldChange` | `pvalue` | `padj` | shift row name into a `gene` column |
| edgeR `topTags()` | row name | `logFC` | `PValue` | `FDR` | rename `FDR` to `padj` |
| DEP / proteoDA | `name` | `_ratio` columns | `_p.val` columns | `_p.adj` columns | wide → long first |
| MaxQuant / Perseus | `Gene names` | `t-test difference` | `-LOG(t-test p-value)` | none | back-transform `-log10 p` first |

## The enrichment frame

The enrichment frame is one row per pathway. `NES` is the signed
effect-size column (positive for up-regulated, negative for
down-regulated); `padj` is the multiple-testing-corrected p; `size` is
the gene-set size used by the test.

| Column | Type | Required? | Notes |
|---|---|---|---|
| `term_col` | character | yes | pathway / term name |
| `nes_col` | numeric | yes | signed; NA arcs are dropped |
| `padj_col` | numeric in [0, 1] | yes | shared name with the volcano frame |
| `size_col` | numeric | only when `magnitude = "size"` | gene-set size |
| `genes_col` | character or list | optional | leading-edge genes |

### Cross-walk

| Upstream | `term_col` | `nes_col` | `padj_col` | `size_col` | `genes_col` (Q4) |
|---|---|---|---|---|---|
| `fgsea::fgseaMultilevel()` | `pathway` | `NES` | `padj` | `size` | `leadingEdge` (list-col, auto) |
| `clusterProfiler::gseGO()` | `Description` or `ID` | `NES` | `p.adjust` | `setSize` | `core_enrichment` ("/" sep, auto) |
| `enrichR::enrichr()` | `Term` | derived from `Odds.Ratio` | `Adjusted.P.value` | none for ORA | `Genes` (";" sep, auto) |
| ORA via clusterProfiler | `Description` | derive from over/under | `p.adjust` | `Count` | `geneID` ("/" sep) |

Pass column overrides when the names don't match — for example with
clusterProfiler GSEA output:

```{r cp, eval = FALSE}
volcano_ring(
  da, gseaResult@result,
  term_col = "Description",
  padj_col = "p.adjust",
  size_col = "setSize",
  genes_col = "core_enrichment",
  genes_sep = "/"
)
```

## Tick-column auto-detect (Q4)

When `genes_col = NULL` (the default), `volcano_ring()` walks four
conventional column names in order:

1. `leading_edge` (`;`-separated string)
2. `leadingEdge` (list-column, fgsea native)
3. `core_enrichment` (`/`-separated string, clusterProfiler GSEA)
4. `Genes` (`;`-separated string, enrichR)

First match wins. A `cli_alert_info()` names the column. No match
silently turns tick lines off — the rings still draw, just without the
radial tick segments inside the volcano.

Override with `genes_col = "my_column"` and, for a string column, a
matching `genes_sep` (defaults to `";"` when the column is a string).

## Renaming strategy

Three patterns work, in increasing order of intrusiveness:

1. **Argument-only renames**: `volcano_ring(..., padj_col = "adj.P.Val")`.
   No frame mutation, good for one-off plots.
2. **Boundary rename**: `names(da)[names(da) == "adj.P.Val"] <- "padj"`.
   Useful when several functions in the pipeline expect the same name.
3. **Built-in column shim**: write a tiny adapter
   `to_evolc(df)` that returns the renamed frame. Worth doing when you
   have a project-specific upstream format you reuse a lot.

When the DA and enrichment sides disagree on the `padj` column name
(e.g. `adj.P.Val` vs `padj`), one of them has to be renamed at the
boundary — `padj_col` is a single shared argument by design.

## NA + edge-case policy

| Case | Behaviour |
|---|---|
| logFC or P.Value NA | row dropped |
| every NES NA | error (cannot colour arcs) |
| pathway NA | row dropped from the ring |
| padj NA in enrichment | filled with 1 |
| NES NA in enrichment | filled with 0 |
| duplicate term | warning + dedup by lowest padj |
| no tick column found | tick lines silently off, one alert |
| p outside [0, 1] | error |

## Upstream tools we consume

The package is a thin plot wrapper around tidy outputs from existing
enrichment tooling:

- fgsea — Korotkevich G et al. 2021, DOI 10.1101/060012
- clusterProfiler — Wu T et al. 2021 *Innovation*
- enrichR — Chen EY et al. 2013 *BMC Bioinformatics*

For methodology when *choosing* which enrichment to run, see:

- Xiao Y et al. 2014 *Bioinformatics* 30(6):801-7, PMID 22321699 (π-value)
- Timmons JA et al. 2015 *Genome Biol* 16:186, PMID 26346307 (ORA universe)
- Reimand J et al. 2019 *Nat Protoc* 14:482-517, PMID 30664679 (protocol)
- Wijesooriya K et al. 2022 *PLoS Comput Biol* 18:e1009935, PMID 35263338 (background bias)
