---
title: "Pathway deduplication"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pathway deduplication}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(enrichVolcano)
```

## Why deduplicate

Pathway databases overlap. MSigDB Hallmark has 50 gene sets. Reactome
has more than 1600 in the same release. KEGG, WikiPathways, and
BioCarta all describe glycolysis with slightly different boundaries.
A single fgsea run against three databases will hit "glycolysis"
three times under three names.

This becomes a figure problem. The ring panel in `enrich_volcano()`
shows the top N pathways by `padj`. Without dedup, six of those ten
slots can describe the same biology with different vocabularies. A
reader looking at the ring sees "HALLMARK_GLYCOLYSIS,
REACTOME_GLYCOLYSIS, KEGG_GLYCOLYSIS" and either thinks the result is
overstated or has to mentally collapse the duplicates anyway.

`ev_collapse()` does that collapsing in code. There are two methods:
one based on gene-set overlap, one based on the leading-edge logic
inside `fgsea::collapsePathways`.

## Jaccard overlap (data-independent)

The Jaccard index is the size of the intersection over the size of
the union of two sets:

$$
J(A, B) = \frac{|A \cap B|}{|A \cup B|}
$$

If A has 50 genes and B has 60, and they share 40, then
$J = 40 / 70 \approx 0.57$. By default `ev_collapse()` calls that a
duplicate and keeps the one with the smaller `padj`.

A worked example with two synthetic pathways:

```{r jaccard-manual}
set_a <- c("ENO1", "PKM", "GAPDH", "HK2", "ALDOA", "PFKM",
           "LDHA", "TPI1", "PGAM1", "PGK1")
set_b <- c("ENO1", "PKM", "GAPDH", "HK2", "ALDOA", "PFKM",
           "LDHA", "TPI1", "PGAM1", "ENO2")  # 9 of 10 shared

inter <- length(intersect(set_a, set_b))
uni <- length(union(set_a, set_b))
inter / uni
```

The two glycolysis-style sets share 9 of 11 unique genes;
$J \approx 0.82$. With the default cutoff of 0.5 they would collapse
to one representative.

A note on the inputs. `ev_collapse()` computes Jaccard on the
`leading_edge` column of the enrichment table, not on the full
pathway gene list. The leading edge is the subset of pathway genes
that drove the fgsea signal in your data. Two pathways that share
50% of their full memberships can still have disjoint leading edges
in a given study and survive dedup. Two pathways with low overall
overlap can have identical leading edges and collapse. This makes
the operation data-aware in practice, even though the index itself
is a property of the gene sets.

The `keep_by` argument picks the tie-breaker. Default is `padj`
(smallest wins). `size` keeps the larger gene set, which is useful
when you want to show the broader category. `NES` keeps the strongest
signal in either direction.

In the package code (`R/collapse.R`), the loop is short: for each
cluster of overlapping pathways, sort by `keep_by` and walk the
list, dropping any pathway whose Jaccard with an already-kept member
exceeds the cutoff. The output is deterministic given the inputs.

## fgsea::collapsePathways (data-dependent)

`fgsea` ships its own deduplication: `fgsea::collapsePathways()`. It
takes the original `fgsea` result, the ranked gene statistics, and
the full pathway list. For each pair of significant pathways, it
removes the leading-edge genes of the more-significant pathway from
the rank vector and asks whether the less-significant pathway is
*still* enriched. If not, the less-significant one is dropped because
its signal came entirely from the same genes that drove the first.

This is conceptually different from Jaccard. Two pathways with low
gene overlap can still be flagged as redundant if they enrich on the
same handful of leading-edge genes. Two pathways with high gene
overlap can still survive if their leading edges happen to be
disjoint.

`ev_collapse(method = "collapse_fgsea")` (and `method = "both"`, which
runs Jaccard first and then collapsePathways) is fully implemented. It
relies on the `ev_pathways` and `ev_stats` attributes that `ev_enrich()`
attaches to its output — the pathway gene sets and the per-contrast
ranked statistics — so pass the `ev_enrich()` result straight through
without dropping those attributes. From the hero function, select it
through the `dedup` list:

```{r collapse-from-hero, eval = FALSE}
p <- enrich_volcano(
  data, contrast = "ctr",
  dedup = list(method = "both", scope = "within_db")
)
```

If the attributes are missing (for example, a hand-built enrichment
table), `ev_collapse()` warns and keeps every pathway rather than
failing. Jaccard remains the default because it is data-independent and
reproducible across studies; reach for collapsePathways when you want a
single study's leading-edge structure to drive the non-redundant set.

## When to pick which

Jaccard suits aggregated multi-database ring plots and any analysis
where reproducibility across studies is the priority. Two pipelines
calling `ev_collapse()` on the same enrichment input with the same
cutoff produce the same kept set.

`collapsePathways` suits a single-database fgsea report where you
want a non-redundant top-line story driven by the actual gene
rankings in your study. The same database analysed in two different
studies can give different non-redundant lists, because the leading
edges differ. That is a feature when reporting one experiment and a
bug when comparing across experiments.

## Scope: within_db vs cross_db

`ev_collapse()` has three scopes.

```{r scope-table, echo = FALSE}
tibble::tribble(
  ~scope,        ~groups_dedup_within,                ~typical_use,
  "within_db",   "contrast + database",               "YvO default; preserves source diversity",
  "cross_db",    "contrast only",                     "Mito default; most aggressive",
  "global",      "none",                              "Cross-contrast, cross-database; rare"
)
```

`within_db` (default) runs Jaccard separately inside each (contrast,
database) bucket. HALLMARK_GLYCOLYSIS and REACTOME_GLYCOLYSIS will
both survive, because they live in different databases. This matches
the YvO 2025 pipeline.

`cross_db` runs Jaccard across all databases inside one contrast.
HALLMARK_GLYCOLYSIS and REACTOME_GLYCOLYSIS now compete; the more
significant one wins. This matches the Mito 2026 convention.

`global` ignores both database and contrast. It is the most
aggressive option and rarely the right choice; an Aging pathway and a
Training pathway can collapse even though they describe different
biology.

## Worked example with the YvO subsample

The dedup mechanics are the same whichever fixture you use. The
package install bundles the YvO results, so the example below uses
that:

```{r yvo-prep}
fixture <- system.file(
  "extdata", "examples", "yvo_tidy.rds",
  package = "enrichVolcano"
)
if (!nzchar(fixture)) {
  fixture <- "../tests/testthat/fixtures/yvo_tidy.rds"
}
res <- readRDS(fixture)
res <- res[!is.na(res$gene) & !is.na(res$contrast), ]
res <- pi_score(res, variant = "eq2")
res <- adjust_p(res, method = "BH")
```

Build two mock databases with a deliberate overlap between
`HALLMARK_GLYCOLYSIS` (in the first DB) and `REACTOME_GLYCOLYSIS` (in
the second DB):

```{r yvo-mock-db}
set.seed(7)
genes <- unique(res$gene)
core <- sample(genes, 30)

hallmark_mock <- list(
  HALLMARK_GLYCOLYSIS = c(core, sample(setdiff(genes, core), 10)),
  HALLMARK_OXPHOS     = sample(genes, 25),
  HALLMARK_APOPTOSIS  = sample(genes, 20)
)
reactome_mock <- list(
  REACTOME_GLYCOLYSIS = c(core, sample(setdiff(genes, core), 5)),
  REACTOME_TCA        = sample(genes, 18),
  REACTOME_BETAOX     = sample(genes, 22)
)
```

`HALLMARK_GLYCOLYSIS` and `REACTOME_GLYCOLYSIS` share the 30 `core`
genes plus a few extras; their pairwise Jaccard sits well above 0.5.

Run enrichment and collapse:

```{r yvo-enrich}
e <- ev_enrich(
  res,
  contrast = "Training_Old",
  databases = list(hallmark_mock = hallmark_mock,
                   reactome_mock = reactome_mock),
  enrich_mode = "fgsea",
  rank_by = "logFC",
  min_size = 5, max_size = 200
)
e[, c("database", "pathway", "padj", "NES")]
```

Apply Jaccard dedup at two scopes:

```{r yvo-collapse}
d_within <- ev_collapse(e, method = "jaccard", cutoff = 0.5,
                        scope = "within_db")
d_cross  <- ev_collapse(e, method = "jaccard", cutoff = 0.5,
                        scope = "cross_db")
c(within_db_kept = sum(d_within$dedup_kept),
  cross_db_kept  = sum(d_cross$dedup_kept))
```

With `within_db`, every pathway is kept: the two glycolysis sets live
in different databases and never get compared. With `cross_db`, the
two glycolysis sets compete and the higher-`padj` one is dropped.

Inspect the `dedup_kept` flag directly:

```{r yvo-inspect}
d_cross[, c("database", "pathway", "padj", "dedup_kept")]
```

`ring_plot()` and `enrich_volcano()` both honour `dedup_kept` and only
plot rows where it is `TRUE`. If you want to compare before and after
in the same figure, run `ev_enrich()` once, then call `ev_collapse()`
with two different scopes and pass each result to `ring_plot()`
separately.

Next: how to pick the right gene-set database for your question. See
`vignette("databases")`.

## References

Korotkevich G, Sukhov V, Sergushichev A. Fast gene set enrichment
analysis. *bioRxiv* 060012 (2019). (fgsea package)

Subramanian A, Tamayo P, Mootha VK, et al. Gene set enrichment
analysis: a knowledge-based approach for interpreting genome-wide
expression profiles. *PNAS* 102(43):15545-15550 (2005).
