---
title: "Pi-score and p-value adjustment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pi-score and p-value adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(enrichVolcano)
```

## The problem with ranking by p-value alone

A p-value answers a single question: how surprising is this effect
under the null. It says nothing about how big the effect is. In
proteomics this matters because two proteins can carry the same
p-value while telling different biological stories.

Consider two rows from a hypothetical limma output:

| gene | logFC | P.Value | -log10(P) |
|------|-------|---------|-----------|
| A    | 0.10  | 1e-6    | 6.0       |
| B    | 2.50  | 1e-3    | 3.0       |

By raw p-value, A wins by three orders of magnitude. A 10% change in
abundance, however, is rarely the headline story in a muscle biopsy.
B has 5.6-fold higher absolute log fold change and a p-value that
would still pass most cut-offs. Ranking by `-log10(p)` alone puts A on
top; ranking by `|logFC|` alone puts B on top. Neither ranking is
right on its own.

Xiao Y et al. (2014, *Bioinformatics* 30(6):801) proposed combining
the two into a single ranking score. They called it pi-score and gave
two equivalent forms. This package implements both.

## Pi-score: two variants

### Equation 2 (default)

$$
\pi_{\text{Eq2}} = P^{|\log_2 \text{FC}|}
$$

Range: $[0, 1]$. Smaller pi means stronger evidence (p is small *and*
logFC is large). The convention threshold is 0.05, mirroring the
familiar p-value threshold.

This is the form used in the YvO, CvH, HRvLR, and Mito pipelines that
ship as fixtures with this package. `pi_score()` defaults to it
because the [0, 1] range plays well with downstream cut-offs that
already expect a probability-like quantity.

```{r show-eq2}
example <- tibble::tibble(
  gene    = c("A", "B"),
  logFC   = c(0.10, 2.50),
  P.Value = c(1e-6, 1e-3)
)
pi_score(example, variant = "eq2")
```

By Eq.2, B has the smaller pi (better) even though its raw p is
larger. The exponentiation by `|logFC|` shrinks A's tiny p when the
effect is small.

### Equation 1 (alternative)

$$
\pi_{\text{Eq1}} = |\log_2 \text{FC}| \times -\log_{10} P
$$

Range: $[0, \infty)$. Larger pi means stronger evidence. Same paper,
same authors, just the other equation in the original text.

```{r show-eq1}
pi_score(example, variant = "eq1")
```

Eq.1 gives `0.10 * 6 = 0.6` for A and `2.50 * 3 = 7.5` for B. B again
beats A, by a wider margin than Eq.2 gives.

### Worked numerical example

Drop the formula into a calculator with logFC = 2 and P = 0.01:

* Eq.2: $0.01^{|2|} = 0.0001$
* Eq.1: $|2| \times -\log_{10}(0.01) = 2 \times 2 = 4$

Now nudge the effect down to logFC = 0.5 with the same p:

* Eq.2: $0.01^{|0.5|} = 0.1$
* Eq.1: $|0.5| \times -\log_{10}(0.01) = 0.5 \times 2 = 1$

Cutting the effect in half multiplies Eq.2 by 1000 and divides Eq.1
by 4. Both penalise small effects, but Eq.2 is more aggressive.

## Choosing a variant

Use Eq.2 when downstream code expects a [0, 1] scale (significance
flags, FDR-like cut-offs, sortable as a p-value). Use Eq.1 when you
need to compare against a paper that reports unbounded pi values. For
fgsea ranking, the two variants rank-order genes the same way as long
as every raw p is below 1 (which is always true in practice). For
ORA, only the [0, 1] form has a natural significance threshold.

## Pi-score is for the volcano, never for GSEA ranking

Both pi-score variants ignore the *sign* of logFC: Eq.2 is bounded in
$[0, 1]$ and Eq.1 is non-negative, so up- and down-regulation collapse
onto the same axis. That is exactly what you want for the volcano
y-axis and the up/down significance call, but it is wrong for ranking
a GSEA.

Gene-set enrichment analysis needs a **signed** ranking statistic so
that up-regulated genes sit at one end of the ranked list and
down-regulated genes at the other (Subramanian et al. 2005; Reimand et
al. 2019). Feeding an unsigned score puts every strong gene at the same
end, collapses the running-sum to one tail, and produces near-empty
rings. `ev_enrich()` therefore ranks by a signed statistic by default:

$$
\text{signed\_p} = \operatorname{sign}(\log_2 \text{FC}) \times -\log_{10} P
$$

This is the package default (`rank_by = "signed_p"`), and the hero
`enrich_volcano()` hard-codes it. A raw moderated *t*-statistic
(`rank_by = "t"`) or `rank_by = "logFC"` are equally valid signed
choices. To guard the invariant, `ev_enrich()` **rejects** any attempt
to rank by an unsigned pi-score column:

```{r reject-pi, error = TRUE}
data <- pi_score(example, variant = "eq2")
data$contrast <- "ctr"
# this is the wrong thing to do, and the package stops you:
ev_enrich(data, contrast = "ctr",
          databases = list(toy = list(SET = c("A", "B"))),
          rank_by = "pi_eq2")
```

In short: pi-score answers "is this protein significant *and* large?"
for the volcano; the signed statistic answers "which direction and how
strongly?" for the ring. They are different questions and must use
different scores.

## Multiple-testing correction overview

`adjust_p()` accepts four methods. They differ in what they
control and how much power you keep.

### Benjamini-Hochberg (default)

BH controls the false discovery rate: the expected proportion of
false positives among rejected hypotheses. It is the proteomics
field default because it gives more power than Bonferroni at a
biologically defensible error rate. `stats::p.adjust(p, "BH")` is the
implementation.

### Bonferroni

Bonferroni controls the family-wise error rate: the probability of
*any* false positive. Multiplies each p by the number of tests. With
3000 proteins this is severe; a raw p of 1e-5 ends up at 0.03. Use
this when you need to confirm a small set of pre-selected biomarkers,
not for screening.

### Storey q-value

The q-value method (Storey 2003, *JRSS-B* 64:479) estimates the
proportion of true null hypotheses ($\pi_0$) and weights BH by it.
When most proteins are truly null, $\hat\pi_0$ is close to 1 and the
result matches BH. When many proteins really do change (a strong
treatment), $\hat\pi_0$ is below 1 and the q-value beats BH on power.
Requires the `qvalue` package from Bioconductor.

### IHW

Independent Hypothesis Weighting (Ignatiadis et al. 2016, *Nature
Methods* 13:577) is BH with informative weights. You supply a
covariate that is independent of the p-value under the null but
correlated with power under the alternative. In proteomics, peptide
count per protein is a good covariate: more peptides means a more
precise estimate, so a small p with high peptide count is more
trustworthy than a small p with one peptide. `adjust_p()` accepts
`covariate = "peptides"` (or any numeric column) and calls
`IHW::ihw()`.

### Choosing in practice

Most proteomics papers report BH. q-value is a free-lunch upgrade
when you have a treatment with global effects. IHW is the best choice
when you have a per-protein covariate that predicts power; otherwise
it reduces to BH.

## Worked example with the YvO subsample

```{r yvo-load}
fixture <- system.file(
  "extdata", "examples", "yvo_tidy.rds",
  package = "enrichVolcano"
)
if (!nzchar(fixture)) {
  fixture <- "../tests/testthat/fixtures/yvo_tidy.rds"
}
yvo <- readRDS(fixture)
yvo <- yvo[!is.na(yvo$gene) & !is.na(yvo$contrast), ]
nrow(yvo)
```

Compute pi-scores both ways and look at the rank correlation within
the Aging contrast:

```{r yvo-pi}
yvo <- pi_score(yvo, variant = "eq2")
yvo <- pi_score(yvo, variant = "eq1")
aging <- yvo[yvo$contrast == "Aging", ]
cor(rank(aging$pi_eq2), rank(-aging$pi_eq1))
```

The correlation is essentially 1: -pi_eq1 and pi_eq2 rank rows
identically. The two are interchangeable for ordering; they differ
only in scale.

Now compare BH and Bonferroni at $\alpha = 0.05$:

```{r yvo-adjust}
yvo_bh   <- adjust_p(yvo, method = "BH")
yvo_bonf <- adjust_p(yvo, method = "bonferroni")
sig_bh   <- sum(yvo_bh$adj.P.Val   < 0.05, na.rm = TRUE)
sig_bonf <- sum(yvo_bonf$adj.P.Val < 0.05, na.rm = TRUE)
c(BH = sig_bh, Bonferroni = sig_bonf)
```

BH calls roughly an order of magnitude more proteins significant than
Bonferroni, at the cost of allowing a few false positives among the
rejected nulls. For a screening workflow that feeds into pathway
enrichment, that trade-off is the right one.

`adjust_p()` adjusts within each contrast by default
(`by_contrast = TRUE`). This matches the proteoDA convention: a
training response and an aging effect are independent families and
should not share an error budget.

Next: how to deduplicate redundant enrichment hits. See
`vignette("pathway-dedup")`.

## References

Xiao Y, Hsiao TH, Suresh U, Chen HIH, Wu X, Wolf SE, Chen Y. A novel
significance score for gene identification from RNA-seq data.
*Bioinformatics* 30(6):801-807 (2014).

Subramanian A, Tamayo P, Mootha VK, et al. Gene set enrichment
analysis: a knowledge-based approach for interpreting genome-wide
expression profiles. *PNAS* 102(43):15545-15550 (2005).

Reimand J, Isserlin R, Voisin V, et al. Pathway enrichment analysis
and visualization of omics data using g:Profiler, GSEA, Cytoscape and
EnrichmentMap. *Nature Protocols* 14:482-517 (2019).

Benjamini Y, Hochberg Y. Controlling the false discovery rate: a
practical and powerful approach to multiple testing. *JRSS-B*
57(1):289-300 (1995).

Storey JD. The positive false discovery rate: a Bayesian
interpretation and the q-value. *Annals of Statistics* 31(6):2013-2035
(2003).

Ignatiadis N, Klaus B, Zaugg JB, Huber W. Data-driven hypothesis
weighting increases detection power in genome-scale multiple testing.
*Nature Methods* 13(7):577-580 (2016).
